Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-06-18 analysis/memo2_TAS_reviewers.Rmd
revised response to reviewer 2's comments 1, 3.

Subsetting the historical data does not necessarily reduce the degree of incompatibility between the historical strike data and the current trial ($b_hist$ - $b_curr$). As we discuss in the paper, it depends on what one assumes about how much the subsetting criteria (e.g., the salient criteria for "similar" cases) changes the degree of incompatibility. We stress "assumes", because the true values of $b_hist$ and $b_curr$ are not directly observable in actual cases. Accordingly, for any subset of the historical strike data one chooses, our guidance for choosing an $\alpha$ value remains the same: Check for how sensitive the posterior estimate of illegal strike bias is to different values of $\alpha$. The current version of the software prototype does not let users subset the historical data by some characteristic (e.g., date, defendant race). However, in principle, it could be modified to add that kind of feature.

@@

Thank you for this comment.  We have revised the paper accordingly. [TBD]

@@

First, we read this comment to posit that judge is unlikely to find ("seriously entertain") even a _Batson_ prima facie case where "data in the present case indicated the removal rates of protected and unprotected members of the venire were (essentially) equal." However, in some cases, bias can be detected even if equal strikes _rates_.  To illustrate, consider the following hypothetical patterns of strikes by a prosecutor:

@@

Here, the strike rates for the cognizable class and the non-cognizable class are equal (5 out of 10).  However, the 95% credible interval for the posterior estimate of $b$ is excludes zero, thus indicating illegal bias.

Second, we took this comment to posit that a judge is unlikely to find even a _Batson_ prima facie case if the posterior estimate of $b$ with $\alpha = 0$ detects no illegal bias against the cognizable class (e.g. Black prospective jurors), even if the posterior estimate of $b$ given some set of non-zero values of $\alpha$ detects illegal bias.  This might occur, for example, where an attorney is motivated to strike Black prospective jurors because they are Black _only_ in cases with Black defendants; all that attorney's past trials involved Black defendants; but the current trial has a White defendant. In this extreme scenario, we agree that a judge might properly worry about high incompatibility and not find a _Batson_ prima facie case for that reason, even though no one can directly observe or know with complete certainty whether this is a case where $b_curr = 0$ and $b_hist >> 0$. Or the judge may still find enough for a prima face case, but downweight the historical data when ultimately deciding whether illegal bias likely occurred. In the reverse extreme situation, the posterior estimate of $b$ with $\alpha = 0$ detects illegal bias but does not detect illegal bias when $\alpha > 0$. Here, too, we agree that a judge might properly worry about high incompatibility, except that down-weighting the historical data here increases the support for at least enough a _Batson_ prima facie case.

>>>>

Subsetting the historical data does not necessarily reduce the degree of incompatibility between the historical strike data and the current trial ($b_hist$ - $b_curr$). As we discuss in the paper, it depends on what one assumes about how much the subsetting criteria (e.g., the salient criteria for "similar" cases) changes the degree of incompatibility. We stress "assumes", because the true values of $b_hist$ and $b_curr$ are not directly observable in actual cases. Accordingly, for any subset of the historical strike data one chooses, our guidance for choosing an $\alpha$ value remains the same: Check for how sensitive the posterior estimate of illegal strike bias is to different values of $\alpha$. The current version of the software prototype does not let users subset the historical data by some characteristic (e.g., date, defendant race). However, in principle, it could be modified to add that feature.

@@

We have revised the paper to mention on page 1 that for _Batson_ challenges, the "ultimate inquiry" is whether the striking party "was motivated in substantial part by discriminatory intent." [@flowers2019, p. 2241]

The phrase "significant factor" does not appear in _Miller-El v. Dretke_, 545 U.S. 231 (2005). At best, the Court wrote: "Comparing his strike with the treatment of panel members who expressed similar views supports a conclusion that race was significant in determining who was challenged and who was not. _Id._ at 252 (footnote omitted).

@@

First, we read this comment to posit that judge is unlikely to find ("seriously entertain") even a _Batson_ prima facie case where "data in the present case indicated the removal rates of protected and unprotected members of the venire were (essentially) equal." However, in some cases, bias can be detected even if there are equal strike _rates_.  To illustrate, consider the following hypothetical pattern of strikes by a prosecutor in a trial with a strike procedure like the one we model in the paper.

@@

Here, the strike rates for the cognizable class and the non-cognizable class are equal (5 out of 10). However, the 95% credible interval for the posterior estimate of $b$ is excludes zero, thus indicating illegal bias.  Thus, in this case, the judge would err by relying on the equal strike rates to infer no strike bias.

Second, we took this comment to say that a judge is unlikely to find even a _Batson_ prima facie case if the posterior estimate of $b$ with $\alpha = 0$ detects no illegal bias (the credible interval includes zero) against some cognizable class (e.g. Black prospective jurors), even if the posterior estimate of $b$ given some set of non-zero values of $\alpha$ detects illegal bias (the credible interval excludes zero).  This might occur, for example, where an attorney is motivated to strike Black prospective jurors because they are Black _only_ in cases with Black defendants; all that attorney's past trials involved Black defendants; but the current trial has a White defendant. In this extreme scenario, we agree that a judge might properly worry about high incompatibility and not find a _Batson_ prima facie case for that reason, even though no one can directly observe or know with complete certainty whether this is a case where $b_curr = 0$ and $b_hist >> 0$. Or the judge may still find enough for a prima face case, but downweight the historical data when ultimately deciding whether illegal bias likely occurred. In the reverse extreme situation, the posterior estimate of $b$ with $\alpha = 0$ detects illegal bias but does not detect illegal bias when $\alpha > 0$. Here, too, we agree that a judge might properly worry about high incompatibility, except that down-weighting the historical data here increases the support for at least enough a _Batson_ prima facie case.
